Modular analysis of gene expression data with R

### 2009-11-29 isanote/modular.tex
Added Sven's changes, this time properly.

\begin{abstract}
\section{Summary:}
Large sets of data, like expression profile from many samples, require
analytic tools to reduce their complexity. Classical (bi-)clustering
algorithms typically attribute elements (genes, arrays) to disjoint groups
(``clusters''). Yet, in some cases overlapping cluster assignments would suit
the biological reality much better.

The Iterative Signature Algorithm (ISA) was designed to overcome this and
other limitations of standard clustering algorithms. It aims to reduce the
complexity of very large sets of data by decomposing it into so-called
``modules''. In the context of gene expression data these modules consist of
subsets of genes that exhibit a coherent expression profile only over a
subset of microarray experiments. Genes and arrays may be attributed to
multiple modules and the level of required coherence can be varied resulting
in different ``resolutions'' of the modular mapping. Since the ISA does not
rely on the computation of correlation matrices (like many other tools), it
is extremely fast even for very large datsets.

@@

The ISA identifies modules by an iterative procedure. The algorithm starts
from an input seed (corresponding to some set of genes or samples) which are
refined at each iteration by adding and/or removing genes and/or samples
until the process converges to a stable set, which is referred to as a
transcription module. This iterative procedure is typically performed to a
large number of seeds. In the unsupervised approach these seeds are chosen
randomly to sample uniformly the immense search space. We also implemented a
semi-supervised method, to which we refer as ``smart seeding'', where the
seeds a biased to start with certain sets of genes or samples for which
there is prior knowledge.

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{isa2workflow3}
\caption{Work flow of a typical modular analysis study, with the
  \Rpackage{eisa} package (left); and four bicluster visualization
  techniques (right). The input of the \Rpackagescriptional regulation one often combines several
microarray experiments into a single dataset. In such a case it is important
to properly normalize such data and reduce the bias due to constituent
datasets. Several methods address this challenge,
see e.g.~\cite{johnson07} for an algorithm that has a GNU R implementation.
(S: Do we have a standard flag to call their software?)

@@

% Normalization.
\emph{ISA normalization.}
In each iteration of the ISA weighted sums of expression levels over either
genes or samples are computed.  for a number of . Since different genes
typically show different levels of base
expression and variance, it is important to standardize expression
levels to $Z$-scores. The ISA uses two se, the gene
threshold and the sample threshold. ISA modules can be organized into
a directed graph, a module tree, in which there is an edge from module
$A$ to module $B$, if the ISA converges to module $B$ from module $A$,
with the same threshold parameters  that were used to find module
$B$. A module tree provides a hierarchical modular description of a
data set.

>>>>

\begin{abstract}
\section{Summary:}
Large sets of data, like expression profile from many samples, require
analytic tools to reduce their complexity. Classical (bi-)clustering
algorithms typically attribute elements (genes, arrays) to disjoint groups
("clusters"). Yet, in some cases overlapping cluster assignments would suit
the biological reality much better.

The Iterative Signature Algorithm (ISA) was designed to overcome this and
other limitations of standard clustering algorithms. It aims to reduce the
complexity of very large sets of data by decomposing it into so-called
"modules". In the context of gene expression data these modules consist of
subsets of genes that exhibit a coherent expression profile only over a
subset of microarray experiments. Genes and arrays may be attributed to
multiple modules and the level of required coherence can be varied resulting
in different "resolutions" of the modular mapping. Since the ISA does not
rely on the computation of correlation matrices (like many other tools), it
is extremely fast even for very large datsets.

@@

The ISA identifies modules by an iterative procedure. The algorithm starts
from an input seed (corresponding to some set of genes or samples) which are
refined at each iteration by adding and/or removing genes and/or samples
until the process converges to a stable set, which is referred to as a
transcription module. This iterative procedure is typically performed to a
large number of seeds. In the unsupervised approach these seeds are chosen
randomly to sample uniformly the immense search space. We also implemented a
semi-supervised method, to which we refer as "smart seeding", where the
seeds a biased to start with certain sets of genes or samples for which
there is prior knowledge.

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{isa2workflow3}
\caption{Work flow of a typical modular analysis study, with the
  \Rpackage{eisa} package (left); and four bicluster visualization
  techniques (right). The input of the \Rpackage{eisa} pipeline is a
  BioConductor \Rclass{ExpressionSet} object. This is normalized to
  $Z$-scores (1); the ISA iterations are initiated using "smart" or
  random seeds (2); and the biclusters are then merged and filtered
  based on the robustness measure (3). The \Rpackage{eisa} package
  contains tools to perform many GO, KEGG, etc. enrichment calculation
  quickly (4), and several bicluster visualization tools (5).
  Subfigures \subfig{A}, \subfig{B}, \subfig{C} and \subfig{D} were
generated
  using the acute lymphoblastic leukemia data set, see
  \cite{chiaretti04} and the \Rpackage{ALL} R package.
  \subfig{(A)} A condition plot for a simple bicluster. Each sample is
  represented as a bar, its height is its score in the bicluster.
  Samples above the red (none in this case) and samples
  below the green line are part of the biclusters. The coloring
  indicates cases and controls.
  \subfig{(B)} A module tree. Each rectangle is a bicluster; see the
  definition of the edges in the text. The modules are colored
  according to their Gene Ontology enrichment $p$-values, the codes of
  the enriched GO categories are shown in the top-left corner of the
  rectangles. The top-right corner shows the number of genes and
  conditions in the bicluster. The gene thresholds used for finding
  the modules is shown on the horizontal axes.
  \subfig{(C)} Heatmap for a single module, showing correlated
  expression for the genes and samples. The red lines are the gene and
  sample scores.
  \subfig{(D)} Profile plot, for a single module. On the top plot the
  expression of all samples is plotted, the samples of the module
  with black. The red and green lines show the mean of the two
  groups. The bottom plot is the same for the genes of the module.
}
\label{fig:workflow}
\end{figure*}

A typical modular analysis for gene expression data includes the following
steps:

% Batch correction
\emph{Batch correction.}
In order study the global organization of a transcription program including
many aspects of transcriptional regulation one often combines several
microarray experiments into a single dataset. In such a case it is important
to properly normalize such data and reduce the bias due to constituent
datasets. Several methods address this challenge,
see e.g.~\cite{johnson07} for an algorithm that has a GNU R implementation.
(S: Do we have a standard flag to call their software?)

@@

% Normalization.
\emph{ISA normalization.}
In each iteration of the ISA weighted sums of expression levels over either
genes or samples are computed.  for a number of . Since different genes
typically show different levels of base
expression and variance, it is important to standardize expression
levels to $Z$-scores. The ISA uses two sets of $Z$-scores, one
calculated for each gene across all samples, the other for each sample
across all genes.

% Random and smart seeding.
\emph{Random and smart seeding.}
As mentioned previously, ISA can be performed with random or smart
seeds, based on the application.

% The ISA iteration.
\emph{The ISA iteration.}
This step is essentially performing the ISA and finding the biclusters
from the starting seeds.

% Merging the modules.
\emph{Merging the modules.}
It is possible that more seeds converge to the same, or very similar
biclusters. This step eliminates such duplicates.

% The robustness measure and filtering the modules.
\emph{Robustness of biclusters.}
To access the significance of a bicluster, we designed a robustness
measure, that can be used to filter out spurious modules. This is done
based on performing ISA on the scrambled input data.

% Module trees.
\emph{Module trees.}
The ISA works with two stringency threshold parameters, the gene
threshold and the sample threshold. ISA modules can be organized into
a directed graph, a module tree, in which there is an edge from module
$A$ to module $B$, if the ISA converges to module $B$ from module $A$,
with the same threshold parameters  that were used to find module
$B$. A module tree provides a hierarchical modular description of a
data set.
